A conditional permutation-based approach to test confounder and center effects in machine learning models

### 2021-10-25 manuscript.tex

The 'partial' and 'full' confound tests, as implemented in version 0.13.0 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.

All analysis code used for the simulations is available on github\footnote{https://github.com/spisakt/mlconfound_manuscript}.

@@

'Partial' and 'full' confound testing was performed with default parameters (10000 permutations, 50 Markov-chain Monte Carlo steps) as implemented in version 0.9.0 of the package \emph{'mlconfound'}. Unconditional dependence across the involved variables was investigated with conventional permutation test on the $R^2$ values, with 1000 permutations.

@@

Analysis code is available as jupyter notebook\footnote{\href{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb}{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb}} in the github repository of the 'mlconfound' package.

@@

The resulting functional connectivity estimates were considered as features either (i) in their raw form or after applying (ii) feature regression \citep{rao2017predictive} or (iii) COMBAT \citep{johnson2007adjusting, fortin2018harmonization}.
The investigated confounder variables were 'imaging center' and in-scanner motion, as measured by the mean framewise displacement of \cite{power2014methods}).
Mean FD was non-linearly transformed to normal distribution with the quantile transformation \citep{beasley2009rank} as implemented in \emph{scikit-learn} \citep{pedregosa2011scikit} (see Supplementary Figure \ref{fig:abide-hist} for details).

As COMBAT is not able to handle continuous variables (since it was primarily designed to remove "batch-effects"), motion was binned into 10 groups, based on equdistant data quantiles ranging from 0 to 1.

@@

Analysis code is available as jupyter notebook\footnote{\href{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_abide.ipynb}{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_abide.ipynb}} in the github repository of the 'mlconfound' package.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/sim_normal.eps}
  \caption{\textbf{Type-I error control and power of the conditional permutation-based confound testing for predictive modelling in normally distributed simulated data}. \\
  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette at the bottom) in various simulations settings (100 simulations per tile) with different covariance between the target and the confounder (given as $R^2_{y,c}$ on the vertical axes of the heatmaps), for different sample sizes (N, horizontal axes).
  Rows and columns correspond to simulations with different signal-in-prediction (y+c) and confound-to-signal values, respectively. The first row and the first column contain simulations under H0 (no confounder bias) and separated form the simulations under the alternative hypothesis with dashed line. Positive rates for the simulations under the null and alternative hypotheses can be interpreted as type I error rate and statistical power, respectively. The higher confidence limit for the type I error rates is 0.11 for each tile, with 5 tiles expected to pass it form all H0 tiles.
  The actual weights $w_{Y}$, $w_{C}$ and $w_{e}$ are depicted at the bars on top of each heatmap and color coded as shown in the legend at the bottom. Above each heatmap, the strengths of the associations of $\hat{y}$ to $Y$ and $C$, respectively, are given with $R^2$ values averaged over all simulation cases of the heatmap.
  }
  \label{fig:sim-normal}
\end{figure}

The conditional permutation based 'partial' and 'full' confound tests for predictive modelling have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.

@@

  \includegraphics[width=0.5\paperwidth]{fig/sim_non-norm.eps}
  \caption{\textbf{Robustness of conditional permutation based confound testing to non-normality and non-linearity.} \\
  Simulations included variables with five different degrees of non-normality (top panel), as introduced with various $\delta$ and $\epsilon$ values of the \emph{sinh-arcsinh} transformation (yellow: normally distributed). Fisher's kurtosis and skewness is given for each distribution. False and true positive rates in the simulations under H0 and H1, respectively, for each investigated sample size (N), are depicted by barplots for both linear and sigmoid dependency structure. Upper 95\% binomial confidence interval corresponding to $alpha=0.05$ is shown with a vertical dashed line.}
  \label{fig:sim-non-normal}
\end{figure}

>>>>

The 'partial' and 'full' confound tests, as implemented in version 0.20.0 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.

All analysis code used for the simulations is available on github\footnote{\href{https://github.com/pni-lab/mlconfound-manuscript/tree/main/simulated}{https://github.com/pni-lab/mlconfound-manuscript/tree/main/simulated}}.

@@

'Partial' and 'full' confound testing was performed with 10000 permutations and 50 Markov-chain Monte Carlo steps, as implemented in version 0.20.0 of the package \emph{'mlconfound'}. Unconditional dependence across the involved variables was investigated with conventional permutation test on the $R^2$ values, with 1000 permutations.

All empirical analyses are available as jupyter notebooks on github\footnote{\href{https://github.com/pni-lab/mlconfound-manuscript/tree/main/empirical}{https://github.com/pni-lab/mlconfound-manuscript/tree/main/empirical}}.

@@

@@

The resulting functional connectivity estimates were considered as features either (i) in their raw form or after applying (ii) feature regression \citep{rao2017predictive} or (iii) COMBAT \citep{johnson2007adjusting, fortin2018harmonization}.
The investigated confounder variables were 'imaging center' and 'in-scanner motion', as measured by the mean framewise displacement of \cite{power2014methods})
Mean FD was non-linearly transformed to normal distribution with the quantile transformation \citep{beasley2009rank} as implemented in \emph{scikit-learn} \citep{pedregosa2011scikit} (see Supplementary Figure \ref{fig:abide-hist} for details).

As COMBAT is not able to handle continuous variables (since it was primarily designed to remove "batch-effects"), motion was binned into 10 groups, based on equidistant data quantiles ranging from 0 to 1.

@@

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/sim_normal.eps}
  \caption{\textbf{Type-I error control and power of the conditional permutation-based confound testing for predictive modelling in normally distributed simulated data}. \\
  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette at the bottom) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\yhat}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\yhat}$ (rows) and for different sample sizes (N, columns). Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%. First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounder bias).
  Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively. The higher 95\% confidence limit for the type I error rates is 0.11 for each tile.
  }
  \label{fig:sim-normal}
\end{figure}

The conditional permutation based \emph{partial} and \emph{full confounder tests} for predictive modelling have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.

@@

  \includegraphics[width=0.5\paperwidth]{fig/sim_non-norm.eps}
  \caption{\textbf{Robustness of conditional permutation based confound testing to non-normality and non-linearity.} \\
  Simulations included variables with five different degrees of non-normality (top panel), as introduced with various $\delta$ and $\epsilon$ values of the \emph{sinh-arcsinh} transformation (yellow: normally distributed). Fisher's kurtosis and skewness is given for each distribution. False and true positive rates in the simulations under H0 and H1, respectively, for each investigated sample size (N), are depicted by barplots for both linear and sigmoid dependency structure. Upper 95\% binomial confidence limit corresponding to $alpha=0.05$ is shown with a vertical dashed line.}
  \label{fig:sim-non-normal}
\end{figure}
